---
title: "Getting your blank maps"
output: 
    html_document:
      df_print: tibble
      toc: yes
      toc_float: yes
---

We need to have "blank maps" to color in as thematic maps, and optionally spatial data (points, lines) to add. Two sources:

1. Dowloaded file [focus here].
2. Provided through an R-package [see [other workshop](https://workshop.mhermans.net/thematic-maps-r/01_get_maps.html).

Various formats for spatial data possible: Shapefile, GeoJSON, KML, GPX-tracks, etc.

Recommended: Use `st_read()` function from the `sf` library, as it returns a spatial feature object and is based on [GDAL](https://gdal.org/): hard to find a format not covered.

Points of attention:

* We want sf-objects. Frequently already in the right format, if not convert using [`st_as_sf()`](https://r-spatial.github.io/sf/reference/st_as_sf.html).
* Pay attention to included identifier for merging.

```{r, message=FALSE, warning=FALSE}
# load general packages used below
library(dplyr)   # tidyverse datamanipulation library
library(sf)      # tidyverse spatial features library
library(janitor) # utility function clean_names()
library(mapview)
library(tmap)
library(here)
```

# Example: load railway tracks

```{r, message=FALSE, warning=FALSE, results = 'hide'}
raillines <- st_read(here('data/source/census_1851_raillines/1851EngWalesScotRail_Lines.shp'))
```

```{r}
mapview(raillines)
```

```{r}
qtm(raillines)
```

# Example: load cencus districts

```{r, message=FALSE, warning=FALSE, results = 'hide'}
districts_spatial <- st_read(here('data/source/census_1851_districts/1851EngWalesRegistrationDistrict.shp'))
```


```{r, message=FALSE, warning=FALSE}
mapview(districts_spatial)
```

```{r, message=FALSE, warning=FALSE}
qtm(districts_spatial)
```

