---
title: "Getting your blank maps"
output: 
    html_document:
      df_print: tibble
      toc: yes
      toc_float: yes
---

We need to have "blank maps" to color in as thematic maps, and optionally spatial data (points, lines) to add. Two sources:

1. Dowloaded file [focus here].
2. Provided through an R-package [see [other workshop](https://workshop.mhermans.net/thematic-maps-r/01_get_maps.html).

Various formats for spatial data possible: Shapefile, GeoJSON, KML, GPX-tracks, etc. Downloading or receiving a Shapefile will likely be the most common scenario for historians.

Recommended: Use `st_read()` function from the `sf` library, as it returns a spatial feature object and is based on [GDAL](https://gdal.org/): hard to find a format that is not covered.

Points of attention:

* For further manipulation we want objects supported by the `sf`-library. When loading spatial data through `st_read()`, it is in the correct format, if another route is needed spatial objects can be converted with [`st_as_sf()`](https://r-spatial.github.io/sf/reference/st_as_sf.html).
* Pay attention to the included identifier in the spatial dataset (NIS-code, census-code, etc.) for merging other data in later steps.

```{r, message=FALSE, warning=FALSE}
# load general packages used below
library(dplyr)   # tidyverse datamanipulation library
library(sf)      # tidyverse spatial features library
library(mapview) # library for interactive exploration of spatial data
library(tmap)    # library for thematic maps
library(here)    # optional helper library for loading data without absolute paths
```

# Example: load railway tracks

```{r, message=FALSE, warning=FALSE, results = 'hide'}
raillines <- st_read(here('data/source/census_1851_raillines/1851EngWalesScotRail_Lines.shp'))
```

```{r}
# quick interactive look using mapview() from mapview library
mapview(raillines)
```

```{r}
# quick static thematic map using qtm() from tmap library
qtm(raillines)
```

# Example: load cencus districts

```{r, message=FALSE, warning=FALSE, results = 'hide'}
districts_spatial <- st_read(here('data/source/census_1851_districts/1851EngWalesRegistrationDistrict.shp'))
```


```{r, message=FALSE, warning=FALSE}
mapview(districts_spatial)
```

```{r, message=FALSE, warning=FALSE}
qtm(districts_spatial)
```

