---
title: "Adding data"
output: 
    html_document:
      df_print: tibble
      toc: yes
      toc_float: yes
---

We have a spatial dataframe (step 1), now we want to add our data-of-interest to visualise on the thematic map. Two steps:

1. Read in data-of-interst into R.
2. Join data-of-interest with the spatial dataframe.

Crucial for linking: have a variable/column in your that contains the appropriate spatial identifier. E.g. Census-identifer, NIS-code, NUTS-code, etc.

How to read in your data depends on the format, three recommended R-packages should cover mosts posibilities: 

* [readr](https://readr.tidyverse.org/): read/write plain text formats such as CSV, TXT, etc.
* [readxl](https://readxl.tidyverse.org/): read in Excel-files (.xlsx, .xls).
* [haven](https://haven.tidyverse.org/): read/write datasets from SAS, SPSS, Stata.


```{r, message=FALSE, warning=FALSE}
library(here)
library(tidyr)
library(readxl)
library(dplyr)
library(sf)
library(tmap)
```

# Read in data-of-interest: census 1851

```{r, message=FALSE, warning=FALSE}
# census 1851 datafile: 156.300 records x 11 variables
census_data <- read_xlsx(here('data/source/census1851_occupations_count.xlsx'))
```

```{r, message=FALSE, warning=FALSE}
# show records 1 to 10
census_data %>%
  slice(1:10)
```

# Explore data-of-interest

Exploring and manipulating data uses tidyverse "verbs": filter, select, group_by, join, tally, mutate, etc. 

## Occupations & gender

```{r, message=FALSE, warning=FALSE}
census_data %>%
  group_by(sex) %>%
  tally(occupation_count)
```


```{r, message=FALSE, warning=FALSE}
census_data %>%
  group_by(sex, pst_a_label) %>%
  tally(occupation_count) %>%
  spread(sex, n)
```

```{r, message=FALSE, warning=FALSE}
census_data %>%
  group_by(sex, pst_a_label) %>%
  tally(occupation_count) %>%
  spread(sex, n) %>%
  mutate(pct_women = F / ( F + M) )
```

## Which county has the most bakers in 1851?

```{r, message=FALSE, warning=FALSE}
census_data %>%
  filter(occupation_label == 'baker') %>% # select only records with occup 'baker'
  group_by(county_name) %>%
  tally(occupation_count) %>% # count bakers, grouped by county
  arrange(desc(n)) # sort descending on number of bakers
```

## Percentage secondary-sector workers per district

```{r, message=FALSE, warning=FALSE}
districts_data <- census_data %>%
  group_by(district_id, district_name, pst_a_label) %>%
  tally(occupation_count) %>%
  spread(pst_a_label, n)

districts_data <- districts_data %>%
  mutate(total = primary + secondary + tertiary_dealers + tertiary_sellers + tertiary_services_professions + sector_unspecific + no_occupation) %>%
  mutate(pct_secondary = secondary / total)

districts_data
```


# Join spatial data (districts) and data-of interest (occupation-data)

Join spatial and non-spatial dataframes using general dataframe-join functions (online [cheatsheet for types of joins](https://stat545.com/bit001_dplyr-cheatsheet.html)).

```{r, eval=FALSE}
# Example syntax:
library(dplyr)
data <- left_join(spatial_data, data_of_interest, by = "identifier")
data <- left_join(spatial_data, data_of_interest, by = c("spatial_identifier" = "data_identifier"))
```

Example: join spatial data on districts, and calculated data and percentages on occupations.

```{r, message=FALSE, warning=FALSE, results = 'hide'}
# load spatial data on districts
districts_spatial <- st_read(here('data/source/census_1851_districts/1851EngWalesRegistrationDistrict.shp')) %>%
  mutate(CEN1 = as.numeric(as.character(CEN1))) # make sure identifiers are the same type as in the district data
```

```{r, message=FALSE, warning=FALSE}
# add population data on district-level to the spatial data-set
districts <- left_join(districts_spatial, districts_data, by = c('CEN1' = 'district_id'))
```

```{r, message=FALSE, warning=FALSE}
districts
```

# Test: display joined data on a map

```{r, message=FALSE, warning=FALSE}
qtm(districts, fill = 'total')
```

