---
title: "Manipulating (spatial) data"
output: 
    html_document:
        toc: TRUE
        toc_float: TRUE
---

After obtaining (step 2) a single dataframe with spatial data and data-of-interest, we can manipulate this dataframe before plotting on a map and during explorative data-visualisation. Two types of data-manipulation: 

1. spatial data-manipulations: spatial cropping, union, aggregation, etc.
2. 'regular' data-manipulations: combine categories, calculate percentage of population, etc.

... but crucially, using R these manipulations use the **same (tidyverse) syntax and are done within one dataframe**, encouraging data exploration of spatial dimensions together with other data-dimensions (trends, quantities, time-dimensions, etc.).

FYI, instead of doing the 'regular' data-manipulations in R, you can also do this in SAS, Stata, Excel, etc. if you are (currently) more familiar with those, before reading and joining data (previous step).

```{r, message=FALSE, warning=FALSE}
library(here)
library(sf)
library(tmap)
library(dplyr)
library(readxl)
library(mapview)
```


```{r, message=FALSE, warning=FALSE, results = 'hide'}
# (1) load spatial data
raillines <- st_read(here('data/source/census_1851_raillines/1851EngWalesScotRail_Lines.shp'))

districts_spatial <- st_read(here('data/source/census_1851_districts/1851EngWalesRegistrationDistrict.shp')) %>%
  mutate(CEN1 = as.numeric(as.character(CEN1))) # make sure identifiers are the same type

# (2) load and add data-of-interest
districts_data <- read_excel(here('data/census1851_districts_count.xlsx'))
districts <- left_join(districts_spatial, districts_data, by = c('CEN1' = 'district_id'))
```


# Data-manipulation on joint dataset

```{r, message=FALSE, warning=FALSE}
# datamanipulations such as calculating percentages still work on the single, joint dataset
#  e.g. percentage of those employed in professions
districts <- districts %>%
  mutate(pct_prof = tertiary_services_professions / total)
```

```{r, message=FALSE, warning=FALSE}
# Nearly 50% employed in professions in central districts of London
mapview(districts, zcol = 'pct_prof')
```


# Spatial manipulation on joint dataset

Data-exploration: what is R_DIV variable?

```{r, message=FALSE, warning=FALSE}
# get al tally of the R_DIV variable
divisions <- districts %>%
  group_by(R_DIV) %>%
  tally()
```

```{r, message=FALSE, warning=FALSE}
# 11 divisions in the spatial data
divisions
```

Notice: the group-and-tally step has not aggregated data by counting districts per divisions, it has also **merged those districts into new geometry-objects** at the division-level. These are new spatial boundaries, not present in the original data, demonstrating the interchangability of spatial and 'regular' data-operations.

```{r, message=FALSE, warning=FALSE}
# view the newly created/aggregated spatial division boundaries
mapview(divisions)
```

# Spatial / data-manipulation: select districts in Manchester-region

Mousing-over on the interactive map above, we can see that the division around the Manchester-area is called "North Western". This information allows us to filter the districts down to those that are situated in that division.

```{r, message=FALSE, warning=FALSE}
nwestern <- districts %>%
  filter(R_DIV == 'NORTH WESTERN')
```

Notice that a 'regular' data-operation such as `filter()` works as well on subsetting spatial data, as subsetting 'regular' data.

```{r, message=FALSE, warning=FALSE}
# static plot of Manchester districts
qtm(nwestern, fill = 'pct_secondary')
```

```{r, message=FALSE, warning=FALSE}
# interactive plot of Manchester districts
mapview(nwestern, zcol = 'pct_secondary') + raillines
```

# Export interactive map for sharing

```{r}
map_manchester <- mapview(nwestern, zcol = 'pct_secondary') + raillines
mapshot(map_manchester, url = here('output/map_manchester.html'))
```

Generated HTML-file 'map_manchester.html' + supporting files (folder 'map_manchester_files') is in folder 'output'.

# Summary: load, manipulate, interactively visualise, and share in 7 lines

The exported HTML-file of the interactive mapview()-generated map, can be shared with collaborators, put on a project-site, used during a presentation, etc. to furter explore or demonstrated the spatial data in context.

```{r, eval=FALSE}
# (1) load spatial data
raillines <- st_read(here('data/source/census_1851_raillines/1851EngWalesScotRail_Lines.shp'))
districts_spatial <- st_read(here('data/source/census_1851_districts/1851EngWalesRegistrationDistrict.shp')) %>%
  mutate(CEN1 = as.numeric(as.character(CEN1)))

# (2) load and add data-of-interest
districts_data <- read_excel(here('data/census1851_districts_count.xlsx'))
districts <- left_join(districts_spatial, districts_data, by = c('CEN1' = 'district_id'))

# (3) manipulate data (select Manchester area)
nwestern <- districts %>% filter(R_DIV == 'NORTH WESTERN')

# export & share with collaborators, etc.
map_manchester <- mapview(nwestern, zcol = 'pct_secondary') + raillines
mapshot(map_manchester, url = here('output/map_manchester.html'))
```
