---
title: "Plotting thematic maps with tmap"
output: 
    html_document:
        toc: TRUE
        toc_float: TRUE
---

Recommended R packages for plotting :

* [tmap](https://github.com/mtennekes/tmap#tmap-thematic-maps-in-r): special-purpose package for *t*hematic *map*s in R. [static, **focus here**]
* [mapview](https://r-spatial.github.io/mapview/): quick, single-function to interactively view spatial data(frames). [interactive, demonstrated]
* [ggplot](https://ggplot2.tidyverse.org/): most popular package for datavis in R, recent and growing support for maps.  [static, not discussed here]
* [leaflet](https://rstudio.github.io/leaflet/): flexible, general purpose package for interactive maps. [interactive, not discussed here]

The `tmap`-package is recommended for the likely use cases for static maps historians will encounter, get from reviewers, etc. Relatively easy to get a quick thematic map, as demonstrated by `qtm()`, but also extremely large range of options to to build and tweak final map.

Learn more / questions on maps:

* Books: "[An Introduction to R for Spatial Analysis and Mapping](https://books.google.be/books?id=iwJ6DwAAQBAJ) (2018)", "[Geocomputation with R](https://geocompr.robinlovelace.net/]) (full-text)", "[Data Visualization. A practical introduction](https://socviz.co/)", "[Fundamentals of Data Visualization](https://serialmentor.com/dataviz/geospatial-data.html) (full-text).
* [Questions on Stackoverflow](https://stackoverflow.com).
* [Datacamp courses](https://www.datacamp.com/).


`tmap` In an nutshell:

* Quick 'n dirty thematic maps: **Q**uick **t**hematic **m**ap: `qtm()`
* Build map in layers with lots and lots of control and options: `tm_shape()`, `tm_fill()`, `tm_borders()`, `tm_lines()`, `tm_layout()` etc.
* Cherry-on-top: switch between interactive and static plotting with `tmap_mode('view')` and `tmap_mode('plot')`.

Some resources on tmap:

* Documentation tmap: press F1 with cursor on a tmap-function in Rstudio, and look [online](https://github.com/mtennekes/tmap).
* [Getting started with tmap](https://cran.r-project.org/web/packages/tmap/vignettes/tmap-getstarted.html)
* https://geocompr.robinlovelace.net/adv-map.html
* Background and walkthrough: [Tennekes, M., 2018, tmap: Thematic Maps in R, Journal of Statistical Software, 84(6), 1-39](https://www.jstatsoft.org/article/view/v084i06).


```{r, message=FALSE, warning=FALSE}
library(here)
library(mapview)
library(sf)
library(tmap)
library(dplyr)
library(readxl)
library(scales) # for function percent_format()
```


```{r, message=FALSE, warning=FALSE, results = 'hide'}
# (1) load spatial data
raillines <- st_read(here('data/source/census_1851_raillines/1851EngWalesScotRail_Lines.shp'))

districts_spatial <- st_read(here('data/source/census_1851_districts/1851EngWalesRegistrationDistrict.shp')) %>%
  mutate(CEN1 = as.numeric(as.character(CEN1))) # make sure identifiers are the same type

# (2) load and add data-of-interest
districts_data <- read_excel(here('data/census1851_districts_count.xlsx'))
districts <- left_join(districts_spatial, districts_data, by = c('CEN1' = 'district_id'))
```

# Layered building of thematic maps

`qtm()` is quick and perfect to get the first view, or an initial map to share. But if you need to go futher with tweaking, styling, or more complex maps, building a thematic maps works well with the "layered" approach of tmap for building a thematic map.

Below an non-exhaustive example of building a map in such a way. One important topic not discussed is choise of breaks and the color palette for the fill of a thematic map: this need to be a concious choice for clear presentation and/or to avoid possible bias in presentation. More examples and advise on this [in the previous workshop](https://workshop.mhermans.net/thematic-maps-r/04_plot.html#binning_and_color_palettes).

```{r, message=FALSE, warning=FALSE}
# add spatial (shape) data and a first layer showing the boundaries
tm_shape(districts) +
  tm_borders()
```

```{r, message=FALSE, warning=FALSE}
# add another first layer, the fill of the districts instead of their boundaries
#  tip: sometimes _less_ elements and details such as lines on a map, can make presentation _more_ clear.
tm_shape(districts) +
  tm_fill('pct_secondary')
```






```{r, message=FALSE, warning=FALSE}
# building in different layers, allows for detailed, seperate styling
tm_shape(districts) +
  tm_borders(col = 'white', lwd = 0.5) + # small white border is less distracting then full black on a map
  tm_fill('pct_secondary', title = 'Emploment secondary sector')
```



```{r, message=FALSE, warning=FALSE}
# building in layers allows for stacking different layers with different 
#  spatial features, in this case we add another layer with the railways.
tm_shape(districts) +
  tm_borders(col = 'white', lwd = 0.5) +
  tm_fill('pct_secondary', title = 'Emploment secondary sector') +
  tm_shape(raillines) +
  tm_lines(col = 'grey60')
```

TYhe railway spatial data are not spatial boundaries (polygons) that can be filled, but lines, so we use `tm_lines()`. Similarly, for e.g. adding sizeble points representing quantities you can use `tm_dots()` or `tm_bubbles()`


```{r, warning=FALSE, error=FALSE}
# subset/filter to the Manchester-area
nwestern <- districts %>%
  filter(R_DIV == 'NORTH WESTERN')
  
# tm_text() allows to add a layer with labels for the districts
tm_shape(nwestern) +
  tm_fill(col = 'pct_secondary') +
  tm_text('district_name', size = .5)
```


# Putting it all together

`tmap` contains a wealth of other functions (and their options) to futher tweak and style you map. Here we demonistrate addional options for the legend, adding a source below, etc.

```{r, message=FALSE, warning=FALSE}
map_final <- tm_shape(districts) +
  tm_borders(col = 'white', lwd = 0.5) +
  tm_fill('pct_secondary', title = 'Secondary sector\nemployment', legend.format = percent_format(accuracy = 1), legend.hist = TRUE) +
  tm_shape(raillines) +
  tm_lines(col = 'grey60') +
  tm_layout(main.title = 'Railways & secondary sector employment in 1851', legend.outside = TRUE, frame = FALSE) +
  tm_credits('Source: Cambridge Group for the\nHistory of Population and Social Structure', position = 'right', align = 'right')

map_final
```

```{r, error=FALSE, warning=FALSE}
# save final map
tmap_save(map_final, here('output/england_1851_map_final.png'), width = 1920, height = 1920)
```

# Cherry-on-top: tmap interactive mode

You can switch between static and interactive mode, and re-plot a map to explore interactively. Good for looking at details that might be unclear on the static map, and you need to filter or deal with (eg. [smaller inset map](https://workshop.mhermans.net/thematic-maps-r/questions_answers.html#how_can_i_join_brussels_as_an_inset-map) ).

```{r, warning=FALSE, message=FALSE}
tmap_mode("view")
tm_shape(nwestern) +
  tm_fill(col = 'pct_secondary') +
  tm_text('district_name')
tmap_mode("plot")
```
